MATH 245 Homework 4

### 2024-03-25 PDEs/math245_hw4.Rmd
worked on pdes

---
title: "MATH 245 Homework 4"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-03-22"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

## Question 2  
Find the Fourier-series of $f(x)$. Does the Fourier-series converge (i) pointwise, or (ii)
uniformly?

**(a)**
\begin{equation*} f(x)=
    \begin{cases}
      1-\lvert x \rvert & \lvert x \rvert \leq 1\\
      1 & 1 < \lvert x \rvert \leq \pi\\
    \end{cases}       
\end{equation*}

**(b)**
\begin{equation*} f(x)=\lvert x \rvert=
    \begin{cases}
      -x & -\pi \leq x \leq 0 \\
      x &  0<x\leq \pi \\
    \end{cases}       
\end{equation*}

**(c)**
$$f(x)=x+x^2, \quad -\pi \leq x \leq \pi $$

@@

**(a)** Find the Fourier-sine-series of
\begin{equation*} f(x)=
    \begin{cases}
      1 & 0 < x < \pi /2 \\
      2 &   \pi /2 < x < \pi \\
    \end{cases}       
\end{equation*}

**(b)** Find the Fourier-cosine-series of $f(x)= \lvert \sin{x} \rvert$. Then find
$$\sum^\infty_{k=1} \frac{1}{4k^2-1} $$

**(c)** The Riemann Zeta function is defined for $s > 1$ by
$$ \zeta(s)= \sum^\infty_{n=1} \frac{1}{n^s}$$

By computing the Fourier series of $x^2$ over $-\pi<x<\pi$ and using Parseval’s identity, compute 
$\zeta(4)$.

**(d)** Use the Fourier series in 2c and the pointwise convergence theorem to find $\zeta(2)$. Then find
$$\sum^\infty_{n=1} \frac{1}{(2n-1)^2} $$

## Question 4
Compute the complex Fourier series of the following functions:

**(a)** Compute the complex Fourier series of $f(x) = e^x$ and show that

$$\coth{\pi}= \frac{1}{\pi}+ \frac{2}{\pi}  \left( \frac{1}{1+1^2} +  \frac{1}{1+2^2} +  \frac{1}{1+3^2} + \;...\right) $$

**(b)** Find the complex Fourier series of $xe^{ix}$. Then use your result to find the Fourier series of $x \cos{x}$ and $x \sin{x}$.

## Question 5.

Find the function represented by the new series which is obtained by termwise integration of the following series from $0$ to $x$.

$$ \sum^\infty_{k=1} (-1)^{k+1} \frac{\cos{kx}}{k}=\log{  \left(2 \cos{\left(\frac{x}{2}\right)}\right) }, \quad -\pi<x<\pi$$

>>>>

---
title: "MATH 245 Homework 4"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-03-24"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

## Question 2  
Find the Fourier series of $f(x)$. Does the Fourier-series converge (i) pointwise, or (ii)
uniformly?

### (a)
\begin{equation*} f(x)=
    \begin{cases}
      1-\lvert x \rvert & \lvert x \rvert \leq 1\\
      1 & 1 < \lvert x \rvert \leq \pi\\
    \end{cases}       
\end{equation*}

The Fourier series for a function $f(x)$ on an interval $-l<x<l$ is
$$f(x)=a_0+ \sum_{n=1}^{\infty} \left( a_n \cos{  \left(\frac{n\pi x}{l}\right) }+b_n \sin{ \left(\frac{n\pi x}{l}\right)}\right)$$
where
$$a_0=\frac{1}{2l} \int_{-l}^{l}f(x)dx$$ 
$$ a_n=\frac{1}{l} \int_{-l}^{l}f(x)\cos{  \left(\frac{n\pi x}{l}\right) }dx$$
$$ b_n=\frac{1}{l} \int_{-l}^{l}f(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx$$

Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}f(x)dx \\
    &=\frac{1}{2\pi}\left[ \int_{-\pi}^{-1} 1\;dx + \int_{-1}^{1} (1-\lvert x \rvert) dx+ \int_{1}^{\pi} 1\;dx  \right] \\
    &= \frac{1}{2\pi}\int_{-\pi}^{\pi}1\;dx - \int_{-1}^{1} \lvert x \rvert\, dx \\
    &= 1- 2\int_{0}^{1} x \; dx \\
    &= 1- 1=0
\end{align*}

where we used the evenness of $\lvert x \rvert$ to rewrite $\int_{-1}^{1} \lvert x \rvert \,dx$ as $2\int_{0}^{1} x \; dx$. Thus, $a_0=0$. Now we perform a similar procedure to find $a_n$:

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi} f(x) \cos{(nx)}dx \\
    &= \frac{1}{\pi}\left[ \int_{-\pi}^{-1} \cos{(nx)}\;dx + \int_{-1}^{1} (1-\lvert x \rvert)\cos{(nx)}\; dx+ \int_{1}^{\pi} \cos{(nx)}\;dx  \right] \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi} \cos{(nx)}\;dx - \frac{1}{\pi} \int_{-1}^{1} \lvert x \rvert\cos{(nx)}\; dx
\end{align*}

Since $\sin{(\pi n)}$ and $\sin{(-\pi n)}$ are zero for any integer $n$, the first integral vanishes. We use the symmetry of $\lvert x \rvert \cos{(nx)}$ (both functions are even, and the product of two even functions is even) and rewrite the second integral as $$- \frac{2}{\pi} \int_{0}^{1}  x \cos{(nx)} \; dx $$

We then use integration by parts:
\begin{align*}
&- \frac{2}{\pi} \int_{0}^{1}  x \cos{(nx)}\; dx \\
&= - \frac{2}{n\pi} \left[x \sin{(nx)} \Bigr\rvert_0^1 - \int_{0}^{1}\sin{(nx)}\; dx \right] \\
&= - \frac{2}{n\pi} \left[\sin{(n)}+ \frac{1}{n} \cos{(nx)}\Bigr\rvert_0^1  \right] \\
&= - \frac{2}{n\pi} \left[\sin{(n)}+ \frac{1}{n} \cos{(n)} - \frac{1}{n} \right]
\end{align*}

$$ a_n= - \frac{2 \sin{n}}{n\pi} + \frac{2(\cos{n}-1)}{n^2 \pi} $$
\begin{align*}
b_n &=
\frac{1}{\pi}\left[ \int_{-\pi}^{-1} \sin{(nx)}\;dx + \int_{-1}^{1} (1-\lvert x \rvert)\sin{(nx)}\; dx+ \int_{1}^{\pi} \sin{(nx)}\;dx  \right] \\
    &=  \frac{1}{\pi} \int_{-\pi}^{\pi} \sin{(nx)}\;dx - \frac{1}{\pi} \int_{-1}^{1} \lvert x \rvert\sin{(nx)}\; dx
\end{align*}

But $\sin{x}$ is an odd function and so is $\lvert x \rvert\sin{(nx)}$, so both of these integrals vanish and we find $b_n=0$.

Now inserting the coefficients $a_n$ we found above, we get the Fourier series:
$$f(x)=\frac{2}{\pi} \sum_{n=1}^{\infty}  \left(\frac{\sin{n}}{n}+ \frac{\cos{n}+1}{n^2} \right) \cos{ \left( nx \right)}$$

### (b)
\begin{equation*} f(x)=\lvert x \rvert=
    \begin{cases}
      -x & -\pi \leq x \leq 0 \\
      x &  0<x\leq \pi
    \end{cases}
\end{equation*}

Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}\lvert x \rvert \, dx \\
    &=\frac{1}{\pi}\int_{0}^{\pi} x \;dx \\
    &= \frac{1}{2\pi} x^2 \Bigr\rvert_0^{\pi} = \frac{\pi}{2}
\end{align*}

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{\left(nx\right) }dx \\
    &= \frac{2}{\pi} \int_{0}^{\pi}x\cos{\left(nx\right) }dx \\
    &= \frac{2}{n \pi} x \sin{(nx)} \Bigr\rvert_0^{\pi} - \frac{2}{n \pi}\int_{0}^{\pi}\sin{\left( nx \right) } dx \\
    &= 0 + \frac{2}{n^2 \pi} \cos{(nx)}\Bigr\rvert_0^{\pi} \\
    &= \frac{2}{n^2 \pi} (-1)^{n+1}
\end{align*}

$$b_n = \frac{1}{\pi} \int_{-\pi}^{\pi} f(x) \sin{(nx)}dx $$
But since $\lvert x \rvert$ is even and $\sin{(nx)}$ is odd, the product of the two functions is odd and so the integral vanishes over the symmetric interval.

Then we plug in our formulas for $a_0$ and $a_n$:
$$ f(x) = \frac{\pi}{2} + \frac{2}{\pi} \sum_{n=1}^{\infty} \left[ \frac{1}{n^2} (-1)^{n+1} \cos{(nx)} \right]$$

$$f(x)=x+x^2, \quad -\pi \leq x \leq \pi$$
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}x + x^2 dx \\
    &=\frac{1}{2\pi} \left( \frac{x^2}{2}+ \frac{x^3}{3} \right) \Bigr\rvert_{-\pi}^{\pi} \\
    &= \frac{1}{2\pi} \left( \frac{\pi^2}{2} + \frac{\pi^3}{3} - \frac{\pi^2}{2} + \frac{\pi^3}{3}\right) \\
    &= \frac{1}{2\pi} \left( \frac{2\pi^3}{3} \right) = \frac{\pi^2}{3}
\end{align*}

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{(nx)}dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}dx 
\end{align*}

Applying integration by parts to the first integral,
$$\frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}+\frac{1}{n\pi} \int_{-\pi}^{\pi}\sin{(nx)}$$
$$=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}+\frac{-1}{n^2\pi} \cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}=0+0=0$$

We then use the integration by parts table method demonstrated in class to find that

$$\frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}= \frac{x^2}{n}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi} +  \frac{2x}{n^2}\cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}+\frac{2}{n^3}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi}$$
The first and last terms (with $\sin{(nx)}$) vanish, while the middle term evaluates to $\frac{4 \cos{(n\pi)}}{n^2}$.

$$\frac{4 \cos{(n\pi)}}{n^2}=\frac{4}{n^2} (-1)^{n}$$
Thus, we add this to the result of first integral to find that
$$a_n= \frac{4}{n^2} (-1)^{n} $$
\begin{align*}
b_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\sin{(nx)}dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\sin{(nx)}+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\sin{(nx)}dx 
\end{align*}

But since $x^2$ is even and $\sin{}$ is odd, the product under the second integral is odd functions and thus evaluates to zero over the symmetric interval $-\pi<x<\pi$. Applying integration by parts to the first integral, 
$$b_n=\frac{-1}{n\pi}x\cos{(nx)} \Bigr\rvert_{-\pi}^{\pi}+\frac{1}{n\pi} \int_{-\pi}^{\pi}\cos{(nx)}$$
But since $\cos$ is even and $\cos{\pi}=\cos{-\pi}$, both terms are zero. Thus, $b_n=0$ and

$$f(x)=a_0 + \sum_{n=1}^{\infty}  a_n \cos{  \left(\frac{n\pi x}{l}\right) } $$
$$f(x)=\frac{\pi^2}{3} + 4 \sum_{n=1}^{\infty} \frac{(-1)^{n}\cos{ (nx) }}{n^2}  $$

@@

### a)  
Find the Fourier sine series of
\begin{equation*} f(x)=
    \begin{cases}
      1 & 0 < x < \pi /2 \\
      2 &   \pi /2 < x < \pi \\
    \end{cases}
\end{equation*}

The Fourier sine series of a function $f(x)$ is $$f(x)= \sum_{n=1}^{\infty} A_n \sin{  \left(\frac{n\pi x}{l}\right) }$$
where 
$$A_n = \frac{2}{l} \int_{0}^{l}f(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx $$
$$A_n = \frac{2}{\pi} \int_{0}^{\pi}f(x)\sin{(nx)}dx $$
\begin{align*}
A_n &= \frac{2}{\pi}\left[ \int_{0}^{\pi/2}\sin{(nx)}dx + 2 \int_{\pi/2}^{\pi}\sin{(nx)}dx \right] \\
    &= \frac{2}{\pi}\left[\frac{-1}{n} \cos{(nx)}\Bigr\rvert_{0}^{\pi/2}- \frac{2}{n} \cos{(nx)}\Bigr\rvert_{\pi/2}^{\pi}\right] \\
    &= \frac{2}{n \pi}\left[\cos{\left(\frac{n\pi}{2}\right)}+1-2\cos{(n\pi)} \right]
\end{align*}

 $$f(x)=  \frac{2}{\pi} \sum_{n=1}^{\infty} \frac{1}{n}\left[1+\cos{\left(\frac{n\pi}{2}\right)}-2\cos{(n\pi)} \right] \sin{(nx) }$$

### b)

Find the Fourier-cosine-series of $f(x)= \lvert \sin{x} \rvert$. Then find
$$\sum^\infty_{k=1} \frac{1}{4k^2-1} $$

The Fourier cosine series of a function $f(x)$ is $$f(x)=\frac{1}{2}A_0+  \sum_{n=1}^{\infty} A_n \cos{  \left(\frac{n\pi x}{l}\right) }$$
where 
$$A_n = \frac{2}{l} \int_{0}^{l}f(x)\cos{  \left(\frac{n\pi x}{l}\right) }dx $$
Assuming $l=\pi$, we have
\begin{equation}\label{eqn:3b}
A_n = \frac{2}{\pi} \int_{0}^{\pi}\lvert \sin{x} \rvert\cos{(nx)}dx
\end{equation}

But since $\sin$ is non-negative on the interval $0 \leq x \leq \pi$, we can drop the absolute value bars and use the following trigonometric identity to rewrite the integral: $\sin{\alpha}\cos{\beta} = \frac{1}{2} [ \sin{(\alpha + \beta)} + \sin{(\alpha - \beta)}]$

\begin{align*}
A_n &= \frac{1}{\pi} \int_{0}^{\pi} \sin{((1+n)x)} +\sin{((1-n)x)}dx \\
    &= \frac{1}{\pi}\left[ \frac{-1}{1+n} \cos{((1+n)x)}\Bigr\rvert_{0}^{\pi} -\frac{1}{1-n}\cos{((1-n)x)}\Bigr\rvert_{0}^{\pi} \right] \\
    &= \frac{1}{\pi}\left[ \frac{-1}{1+n} [ (-1)^{1+n}-1] -\frac{1}{1-n} [ (-1)^{1-n}-1]\right] \\
    & = \frac{1}{\pi}  \left[ (-1)^{1+n}-1) \right] \left(\frac{-1}{1+n} -\frac{1}{1-n} \right)
\end{align*}

Thus, we have found a formula for $A_n$,

$$A_n = \frac{-2}{\pi}  \frac{(-1)^{n}+1}{n^2-1}  $$
This formula works for all $n \neq 1$
For $n=0$, 
$$A_0 = \frac{-2}{\pi}  \frac{(-1)^{0}+1}{0-1}=  \frac{4}{\pi}  $$

For $n=1$, we can go back to (\ref{eqn:3b}) and plug in the appropriate value of $n$:
$$A_1 = \frac{1}{\pi} \int_{0}^{\pi} \sin{(2x)}\;dx = \frac{1}{2\pi} \cos{(2x)}\Bigr\rvert_{0}^{\pi} \;dx=0 $$
The Fourier cosine series is therefore
$$f(x)=\frac{2}{\pi}-\frac{2}{\pi}  \sum_{n=2}^{\infty}   \frac{(-1)^{n}+1}{n^2-1}\cos{(nx) }$$

Now, $(-1)^{n}+1$ will be zero for all odd $n$ and two for all even $n$. So substituting $n=2k$, we gwt

$$\lvert \sin{x} \rvert=\frac{2}{\pi}-\frac{4}{\pi}  \sum_{2k=2}^{\infty}   \frac{1}{4k^2-1}\cos{(2kx) }$$

Starting the index at $2k=2$ is equivalent to starting it at $k=1$, so after making that change and evaluating both sides at $x=0$, we can solve for the sum given in the question:
$$0=\frac{2}{\pi}-\frac{4}{\pi}  \sum_{k=1}^{\infty}   \frac{1}{4k^2-1}$$
$$\sum_{k=1}^{\infty}   \frac{1}{4k^2-1}=\frac{1}{2}$$

### c)

The Riemann Zeta function is defined for $s > 1$ by
$$ \zeta(s)= \sum^\infty_{n=1} \frac{1}{n^s}$$
By computing the Fourier series of $x^2$ over $-\pi<x<\pi$ and using Parseval’s identity, compute
$\zeta(4)$.

\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}x^2 dx \\
    &=\frac{1}{2\pi} \left(\frac{x^3}{3}\right) \Bigr\rvert_{-\pi}^{\pi} \\
    &= \frac{1}{2\pi} \left( \frac{\pi^3}{3} - \frac{-\pi^3}{3}\right) \\
    &= \frac{1}{2\pi} \left( \frac{2\pi^3}{3} \right) = \frac{\pi^2}{3}
\end{align*}

$$a_n = \frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}dx$$

But in part (2b), we found that this evaluates to zero.

<!-- **(d)** Use the Fourier series in 2c and the pointwise convergence theorem to find $\zeta(2)$. Then find -->
<!-- $$\sum^\infty_{n=1} \frac{1}{(2n-1)^2} $$ -->

<!-- ## Question 4 -->
<!-- Compute the complex Fourier series of the following functions: -->

<!-- **(a)** Compute the complex Fourier series of $f(x) = e^x$ and show that -->

<!-- $$\coth{\pi}= \frac{1}{\pi}+ \frac{2}{\pi}  \left( \frac{1}{1+1^2} +  \frac{1}{1+2^2} +  \frac{1}{1+3^2} + \;...\right) $$ -->

<!-- **(b)** Find the complex Fourier series of $xe^{ix}$. Then use your result to find the Fourier series of $x \cos{x}$ and $x \sin{x}$. -->

<!-- ## Question 5 -->

<!-- Find the function represented by the new series which is obtained by termwise integration of the following series from $0$ to $x$. -->

<!-- $$ \sum^\infty_{k=1} (-1)^{k+1} \frac{\cos{kx}}{k}=\log{  \left(2 \cos{\left(\frac{x}{2}\right)}\right) }, \quad -\pi<x<\pi$$ -->
